A Framework for Discrete Change

### 2008-11-17 papers/individual/individual.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@242 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%&LaTeX

\documentclass[a4paper,12pt,man,english]{apa} % nobf, doc, apa

@@

\author{Ingmar Visser, Brenda R. J. Jansen, \& Maarten Speekenbrink}

@@

\affiliation{Department of Psychology, University of Amsterdam\\
 Correspondence concerning this article should be addressed to:
 \\
 Ingmar Visser \\
 Department of Psychology, University of Amsterdam \\
 Roetersstraat 15 \\
 1018 WB Amsterdam \\
 phone: +31 (20) 5256723 \\
 fax: +31 (20) 6390279 \\
 email: i.visser@uva.nl}

@@

\abstract{A class of models is developed for measuring and detecting
discrete change in learning and development.  The basic model for
detecting such change is the latent or hidden Markov model.
Traditionally, these models were restricted to categorical, mostly
binary, observed variables, placing severe restrictions on possible
measurement models.  In this paper, the basic model is extended to
include arbitrary distributions for the observed variables, including
multi-variate distributions.  Moreover, there is optional support to
include time-varying predictors.  In effect, this model consists of
mixtures of general linear models with Markovian depencies over time
to model the change process.  In addition, transition parameters can
be made to depend on covariates as well, such that the switching
regime between states depends on characteristics of the individual or
the experimental situation.  The model is illustrated with an example
of participants' learning in the Weather Prediction Task and in the 
Iowa Gambling Task.}

@@

% intro
Discrete change frequently occurs in learning and development: in
learning concepts, in performance on Piagetian tasks, in
discrimination learning and in conditioning.  This chapter is
concerned with detecting the time points of change in (individual)
time series.  We present a framework of dependent mixture models that can
be used to differentiate between gradual and discrete learning events
in individual time series data.  Before presenting the model in formal
terms and providing some illustrations, we first review some examples
in which discrete change is found.

@@

% hysteresis on the balance scale
\citet{Jansen2001} found clear evidence for stage-wise
transitions between Rule 1 and Rule 2 by testing criteria that were
derived from the catastrophe model.  In particular, they found bimodal
test scores and inaccessibility.  The latter means that there are no
in-between strategies: children apply either Rule 1 or Rule 2 and
there is no in-between option.  \citet{Jansen2001} also found
evidence for hysteresis: the phenomenon that switching between
strategies is assymetric.  Children can switch from Rule 1 to Rule 2
and back, but this occurs at different trials.  In particular, if the
distance dimension in the balance scale problems is made more salient
by increasing the distance difference between weights on either side
of the balance scale, children may switch from Rule 1 to Rule 2.  If
subsequently the distance difference is decreased again, children may
switch back to using Rule 1.  Hysteresis is the phenomenon that this
switch back occurs at a different value of the control variable, in
this case the distance difference.

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for sudden
changes in response behavior \citet{Gallistel2004}.  In particular, in
their study, evidence was found for sudden onset of learning: at the
start of the learning experiment, the pigeons did not learn anything
and performance was stable; after a number of trials, learning kicks
in and there are large increases in performance.  The interest here is
in modeling the distribution of onset times: that is, the trials at
which learning suddenly takes off.  A similar interest in process
onset times is found in addiction research.  For example,
\cite{Sher2004} study the age at which children start using alcohol
and how this related to eventual outcomes in terms of addiction.

@@

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
Weather Prediction Task.  The interest in these tasks is to show that
participants develop different strategies over time in responding to
the stimuli and that the transition from one strategy to the next is a
discrete event.  Before providing these illustrations, below we give a
formalization of dependent mixture models and a brief overview of the
DepmixS4 package that was developed to specify and fit such models.

@@

% history of Markov models/LMM applications
Markov models have been used extensively in the social sciences; for
example, in analyzing language learning \cite{Miller1952,Miller1963},
in the analysis of paired associate learning \cite{Wickens1982}.  In
these models, the focus is on survey type data: a few repeated
measurements taken in a large sample.  \citeA{Langeheine1990} discuss
latent Markov models and their use in sociology and political science
(see also McCutcheon, 1987).  Latent transition models have been used
in studying development of math skills \citep{Collins1992} and in
medical applications \citep{Reboussin1998}; \citet{Kaplan2008}
provides an overview of such models, that are called stage-sequential
models in the developmental psychology literature.

@@

% design goals
The \pkg{depmixS4} package was motivated by the fact that Markovian
models are used commonly in the social sciences, but no comprehensive
package was available for fitting such models in R. Common programs
for Markovian models include Panmark \citep{Pol1996}, and for latent
class models Latent Gold \citep{Vermunt2003}.  Those programs are
lacking a number of features that were needed in our own research.  In
particular, \pkg{depmixS4} was designed to meet the following goals:
\begin{enumerate}

@@

Although \pkg{depmixS4} is designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case with
$T=1$ in analyzing cross-sectional data.  In those cases, there are no
time dependencies between observed data, and the model reduces to a
finite mixture model \cite{McLachlan2000}, or a latent class model
\cite{McCutcheon1987}.  Although there are other specialized (R-)
packages to deal with mixture data, one specific feature that we
needed which is not available in other packages is the possibility to
include covariates on the prior probabilities of class membership.

The package is built using S4 classes (object oriented classes in R)
to allow easy extensibility (Chambers, 1998). A depmix model consists 
of the following: 
\begin{enumerate} 
	\item A model for the initial probabilities or prior 
	\item A list of transition models; one model for each row of the 
	transition matrix
	\item A list of response models for each of the states of the model; 
	this is a list of lists in the case of multivariate responses. 
\end{enumerate} 
Each of these are adressed briefly below.

@@

% transition and initial state probs
Each row of the transition matrix and the vector of initial state 
probabilities is modelled as a baseline-category multinomial logistic 
model \cite<see>[chapter 7]{Agresti2002}. This means that covariates 
can be used as predictors in these models. In particular, the model 
for each multinomial is:
\begin{equation} 
	\pi_{i} = \frac{\exp(\alpha_{i}+\mathbf{\beta}^{`}_{i}\vc{x})}
				{1+\sum_{j=1}^{J-1}\exp(\alpha_{j}+\mathbf{\beta}^{`}_{j}\vc{x})},
\end{equation}
where $\pi_{i}$ is the probability (e.g.\ the probability of the
$i$-th initial state); the $\alpha_{i}$ are the category intercepts;
the $\mathbf{\beta}_{i}$ are the category regression coefficients;
$\vc{x}$ is a vector of covariates or predictors; $J$ is the number
of states; in this example, $J$ serves as the baseline-category,
meaning that $\alpha_{J}$ and $\mathbf{\beta}_{J}$ are zero.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  EM is used by
default in unconstrained models, but otherwise, direct optimization is
done using \pkg{Rdonlp2} \cite{Tamura2007,Spellucci2002}, because it
handles general linear (in)equality constraints, and optionally also
non-linear constraints.

@@

% what is the HDT? developmental trends and relevance?
\cite{Crone2004} designed a developmentally appropriate analogue of
the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
schedule although the abolute amounts were redcuced by a factor of 25.
The HDT is a pro-social game inviting the player to assist a hungry
donkey to collect as many apples as possible, by opening one of four
doors.  Again, doors A and B are characterized by a high constant gain
(10 apples), whereas doors C and D deliver a low constant gain (2
apples).  At doors A and C, a loss of 50 apples (A) or 10 apples (C)
is delivered in 50\% of the trials.  For doors B and D, frequency of
loss is only 10\%.  The median loss of doors B and D is 10 and 2,
respectively.  \cite{Crone2004} administered the HDT to children from
four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
that children also fail to consider future consequences.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset \cite{Huizenga2007} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  Most youngest children in the
dataset seem to focus on the dominant dimension in the task, frequency
of loss, resulting in equal preference for doors B and D. Older
participants seem to use a two-dimensional rule where participants
first focus on the frequency of loss and then consider amount of loss,
resulting in a preference for door D. A third very small subgroup
seems to use an integrative rule where participants combine all three
dimensions in the appropriate way.  Participants using the integrative
rule pick cards from doors C and D, which are advantageous in the long
run.

% problematic aspects of standard analysis
Typical analyses of these data use the last 60 trials in a series of
200 trials.  A silent assumption that is made in these analyses is
that behavior has stabilized after 140 trials of learning; this could
very well be wrong and it is highly likely that there are individual
differences in this learning process.

% proposed analysis here
Single participant IGT learning data are analyzed with the aim to
establish 1) at which trial no more learning occurs, i.e., at which
trial behavior has stabilized, and 2) what strategies participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
various participants.  The responses were fitted with multinomial
logistic models for the 4 possible choices that participants make in
this task.  There were no covariates on any of the parameters.  The
only constaint that was imposed on the models' parameters was that
there were designated begin and end states.  This means that the
initial state probabilities were fixed to one for the first state and
to zero for the remaning states.  For the transition matrix this means
that the final state was an absorbing state: the probability of
transitioning out of that state is zero.  This was done to ensure that
there would be a final state in which participants end which provides
the possibility of immediately seeing participants' behavior at the
end of the task.  Models were selected using the Akaike Information
Criterion (AIC; Akaike, 1973).

@@

We chose to analyze two participants that used different strategies at
the end of the task as analyzed in the manner proposed by
\citet{Huizenga2007}.  The first participants' data were best
described by a 4-state model.  This model's transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix the final state is
absorbind.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~ref{tab:igt4} below.

@@

The states have clear interpretations. States 3 and 4 are both
dominated by C responses, and state 4 has a low probability of B
responses. States 1 and 2 are dominated C and D and A and B responses 
respectively. To get a clearer intpretation of these states it is
necessary to consider when they are visited during the learning
process. The viterbi algorithm (REFERENCE) provides us with the
posterior state sequence, i.e., the sequence of states that the
participant is in at each trial. This sequence is depicted in
Figure~\ref{fig:post4}.

@@

At the start of training there are many switches between states 1 and 
2, indicating that behavior is mostly random choices between all four 
categories, albeit with a slight preference for B and D over A and C
choices. This preference indicates a focus on the frequency of loss
associated with each of the doors as B and D have low frequency of
loss. After this, there is a period of stable state 2 behavior
associated with responses A and B. After that there is stable state 3 
behavior, consisting only of C responses, a short transitional period 
and then state 4 behavior consting of 94 \% C choices and some B
choices, i.e., almost optimal behavior. The choice proportions and the
model predicted choice proportions are depicted in
Figure~\ref{fig:igt4}.

@@

The initial preference for B and D choices confirms the theory
expressed in \citet{Huizenga2007} that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices reprensents one of the optimal strategies, which consists of
both C and D responses. Note that both C and D reponses generate equal
profits in the long run.

@@

There are different accounts of probabilistic category learning.
According to instance or exemplar learning theories, participants
learn by storing each encountered cue-outcome pairing.  When presented
with a cue pattern, these exemplars are retrieved from memory, and
weighted according to their similarity to the probe cue pattern, to
form a classification.  According to associative theories,
participants gradually learn by gradually associating the individual
cues (or cue patterns in configural learning) to the outcomes.  In
rule-learning, participants are taken to extract rules by which to
categorize the different cue patterns.  \cite{Gluck2002} proposed a
number of such rules (or strategies).  A main difference between these
is whether responses are based on the presence/absence of a single
cue, or whether responses are based on cue patterns.  Gluck et al.
formulated all strategies in a deterministic and optimal manner (e.g.,
the multi-cue strategy corresponded to giving the optimal response to
each cue pattern).  Meeter et al.  allowed for probabilistic
responding (a small probability of giving the non-optimal response).

Alternative non-strategy based analyses of the WPT
\cite{Lagnado2006,Speekenbrink2008} have estimated response strategies by
logistic regression, allowing the regression coefficients to change
over time.

%associative, rule-based.
Here, we analyze the behavior of a single individual performing the
WPT for 200 trials.  We chose to analyse the ``average'' participant
(the participant with performance closest to the group average) in a
large unpublished dataset.  We let each state be characterized by a
GLM with a Binomial distributed response and logistic link function
(i.e., a logistic regression model).  We are particularly interested
in evidence for strategy switching and whether a DMM can recover a
strategy model in line with \citeA{Gluck2002}.

As we fitted a DMM to the data of a single subject, it was necessary
to place some constraints on the model.  Specifically, we constrain
the state transitions to be in a ``left-right'' format (states can
only proceed to the immediately adjacent state and never back, and
must start in the initial state).  We fitted a single, two and three
state model to the data.  This showed that a two state model was
better than a single or three state model.

\begin{table}
\caption{Estimates for the weather prediction task}
\label{tab:WPT}
\begin{tabular}{lcccccccc} \hline
 & & \multicolumn{1}{c}{1 state} & & \multicolumn{2}{c}{2 state} &&
 \multicolumn{2}{c}{2 state (constr.)} \\ \cline{3-3} \cline{5-6} \cline{8-9}
parameter & & $S_1$ & & $S_1$ & $S_2$ & & $S_1$ & $S_2$ \\ \hline
(intercept) & & -0.69 & & -2.73 & 0.88 & & -1.24 & 0 \\
cue 1 && 1.69 && 2.12 & 1.60 && 1.65 & 1.97 \\
cue 2 && 1.12 && 0.97 & 1.63 && 0 & 1.92 \\
cue 3 && -0.49 && 0.91 & -2.03 && 0 & -1.58 \\
cue 4 && -1.32 && 0.69 & -3.16 && 0 & -2.67 \\ \hline
 & & \multicolumn{1}{c}{AIC=204.47} & & \multicolumn{2}{c}{AIC=187.50} &&
 \multicolumn{2}{c}{AIC=185.24}
\end{tabular}
\end{table}

Investigation of the parameter estimates (see Table~\ref{tab:WPT})
showed that the regression coefficient for the first cue was of much
larger magnitude than that of the other cues.  As such, the first
state seems representative of single cue strategy.  Alternatively, as
all regression coefficients are positive, it could indicate a
``counting'' heuristic, where the propensity of ``sun'' responses
increases when more cues are present (regardless of which cues they
are).  However, in that case, we would expect the regression
coefficients to be of roughly identical magnitude.  The second state
seemed to represent a multi-cue strategy, as all cues had regression
coefficients of reasonable magnitude, with differing directions in
line with the objective cue validities.

To reduce the degrees of freedom, and improve parameter estimates, we
implemented constraints to force state 1 into a single cue strategy
(fixing the coefficients of the remaining three cues to 0) and state 2
in a multi-cue strategy (forcing the intercept to 0).  These
restrictions resulted in a better AIC value of AIC=185.24 (df=7).
Interestingly, the single cue strategy was somewhat different than
described by \citet{Gluck2002}  Parameter estimates indicated relatively
more consistent predictions of ``rain'' in the absence of cue 1
($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun''
in the presence of cue 1 ($Pr(\text{sun}) = 0.60$).  The cue weights
of the multi-cue strategy were in the direction of the optimal
weights.  The Viterbi state sequence indicated that the participant
used the single cue strategy for the first 60 trials, and then
switched to the multi-cue strategy.

>>>>

%&LaTeX
\documentclass[a4paper,12pt,man,english]{apa} % nobf, doc, apa

@@

\twoauthors{Ingmar Visser \& Brenda R. J. Jansen}{Maarten Speekenbrink}

@@

\twoaffiliations{Department of Psychology, University of Amsterdam}{Department of Cognitive, Perceptual and Brain Sciences, University College London}

\note{Correspondence concerning this article should be addressed to:
 \\
 Ingmar Visser \\
 Department of Psychology, University of Amsterdam \\
 Roetersstraat 15 \\
 1018 WB Amsterdam \\
 phone: +31 (20) 5256723 \\
 fax: +31 (20) 6390279 \\
 email: i.visser@uva.nl}

@@

\abstract{A class of models is developed for measuring and detecting
discrete change in learning and development.  The basic model for
detecting such change is the latent or hidden Markov model.
Traditionally, these models were restricted to categorical, mostly
binary, observed variables, placing severe restrictions on possible
measurement models.  In this paper, the basic model is extended to
include arbitrary distributions for the observed variables, including
multi-variate distributions.  Moreover, there is optional support to
include time-varying predictors.  In effect, this model consists of
mixtures of generalized linear models with Markovian dependencies over time
to model the change process.  In addition, transition parameters can
be made to depend on covariates as well, such that the switching
regime between states depends on characteristics of the individual or
the experimental situation.  The model is illustrated with an example
of participants' learning in the Iowa Gambling Task and in the Weather 
Prediction Task.}

@@

% intro
Discrete change frequently occurs in learning and development: in
learning concepts, in performance on Piagetian tasks, in
discrimination learning and in conditioning.  This chapter is
concerned with detecting the nature and time points of change in (individual)
time series.  We present a framework of dependent mixture models that can
be used to differentiate between gradual and discrete learning events
in individual time series data.  Before presenting the model in formal
terms and providing some illustrations, we first review some examples
in which discrete change is found.

@@

% hysteresis on the balance scale
\citet{Jansen2001} found clear evidence for stage-wise
transitions between Rule 1 and Rule 2 by testing criteria that were
derived from the catastrophe model.  In particular, she found bimodal
test scores and inaccessibility.  The latter means that there are no
in-between strategies: children apply either Rule 1 or Rule 2 and
there is no in-between option.  \citet{Jansen2001} also found
evidence for hysteresis: the phenomenon that switching between
strategies is asymmetric.  Children can switch from Rule 1 to Rule 2
and back, but this occurs at different trials.  In particular, if the
distance dimension in the balance scale problems is made more salient
by increasing the distance difference between weights on either side
of the balance scale, children may switch from Rule 1 to Rule 2.  If
subsequently the distance difference is decreased again, children may
switch back to using Rule 1.  Hysteresis is the phenomenon that this
switch back occurs at a different value of the control variable, in
this case the distance difference.

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for sudden
changes in response behavior \cite{Gallistel2004}.  In particular, in
their study, evidence was found for sudden onset of learning: at the
start of the learning experiment, pigeons did not learn anything
and performance was stable; after a number of trials, learning kicked
in and there were large increases in performance.  \citeauthor{Gallistel2004} 
focused on modeling the distribution of onset times: that is, the trials at
which learning suddenly takes off.  A similar interest in process
onset times is found in addiction research.  For example,
\citet{Sher2004} study the age at which children start using alcohol
and how this related to eventual outcomes in terms of addiction.

@@

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
Weather Prediction Task.  The interest in these tasks is to show that
participants develop different strategies over time in responding to
the stimuli and that the transition from one strategy to the next is a
discrete event.  Before providing these illustrations, below we give a
formalization of dependent mixture models and a brief overview of the
\pkg{depmixS4} package that was developed to specify and fit such models.

@@

% history of Markov models/LMM applications
Markov models have been used extensively in the social sciences; for
example, in analyzing language learning \cite{Miller1952,Miller1963} and
in the analysis of paired associate learning \cite{Wickens1982}.  In
these models, the focus is on survey type data: a few repeated
measurements taken in a large sample.  \citeA{Langeheine1990} discuss
latent Markov models and their use in sociology and political science
(see also McCutcheon, 1987).  Latent transition models have been used
in studying development of math skills \citep{Collins1992} and in
medical applications \citep{Reboussin1998}; \citet{Kaplan2008}
provides an overview of such models, that are called stage-sequential
models in the developmental psychology literature.

@@

% design goals
The \pkg{depmixS4} package was motivated by the fact that Markovian
models are used commonly in the social sciences, but no comprehensive
package was available for fitting such models in R. Common programs
for Markovian models include Panmark \citep{Pol1996}, and for latent
class models Latent Gold \citep{Vermunt2003}.  Those programs are
lacking a number of features that were needed in our research.  In
particular, \pkg{depmixS4} was designed to meet the following goals:
\begin{enumerate}

@@

Although \pkg{depmixS4} is designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case with
$T=1$ in analyzing cross-sectional data.  In those cases, there are no
time dependencies between observed data, and the model reduces to a
finite mixture model \cite{McLachlan2000}, or a latent class model
\cite{McCutcheon1987}.  Although there are other specialized (R)
packages to deal with mixture data, these don't allow the inclusion of 
covariates on the prior probabilities of class membership.

\pkg{depmixS4} is built using S4 classes (object oriented classes in R)
to allow easy extensibility (Chambers, 1998). A depmix model consists 
of the following: 
\begin{enumerate} 
	\item A model for the initial or prior probabilities 
	\item A list of transition models; one model for each row of the 
	transition matrix
	\item A list of response models for each of the states of the model; 
	this is a list of lists in the case of multivariate responses. 
\end{enumerate} 
Each of these are adressed briefly below.

@@

% transition and initial state probs
By default, each row of the transition matrix and the vector of initial state 
probabilities is modelled as a baseline-category multinomial logistic 
model \cite<see>[chapter 7]{Agresti2002}. This means that covariates 
can be used as predictors in these models. In particular, the model 
for each multinomial is:
\begin{equation} 
	\pi_{i} = \frac{\exp(\alpha_{i}+\mathbf{\beta}^{`}_{i}\vc{x})}
				{1+\sum_{j=1}^{J-1}\exp(\alpha_{j}+\mathbf{\beta}^{`}_{j}\vc{x})},
\end{equation}
where $\pi_{i}$ is the probability (e.g.\ the probability of the
$i$-th initial state); the $\alpha_{i}$ are the category intercepts;
the $\mathbf{\beta}_{i}$ are the category regression coefficients;
$\vc{x}$ is a vector of covariates or predictors; $J$ is the number
of states; in this example, $J$ serves as the baseline-category,
meaning that $\alpha_{J}$ and $\mathbf{\beta}_{J}$ are zero.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  EM is used by
default in unconstrained models, but otherwise direct optimization is
done using \pkg{Rdonlp2} \cite{Tamura2007,Spellucci2002}, because it
handles general linear (in)equality constraints, and optionally also
non-linear constraints.

@@

% what is the HDT? developmental trends and relevance?
\citet{Crone2004} designed a developmentally appropriate analogue of
the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
schedule although the absolute amounts were reduced by a factor of 25.
The HDT is a pro-social game inviting the player to assist a hungry
donkey to collect as many apples as possible, by opening one of four
doors.  Again, doors A and B are characterized by a high constant gain
(10 apples), whereas doors C and D deliver a low constant gain (2
apples).  At doors A and C, a loss of 50 apples (A) or 10 apples (C)
is delivered in 50\% of the trials.  For doors B and D, frequency of
loss is only 10\%.  The median loss of doors B and D is 10 and 2,
respectively.  \citet{Crone2004} administered the HDT to children from
four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
that children also fail to consider future consequences.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset \cite{Huizenga2007} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  The youngest children in the
dataset mostly seem to focus on the dominant dimension in the task, frequency
of loss, resulting in equal preference for doors B and D. Older
participants seem to use a two-dimensional rule where participants
first focus on the frequency of loss and then consider amount of loss,
resulting in a preference for door D. A third very small subgroup
seems to use an integrative rule where participants combine all three
dimensions in the appropriate way.  Participants using the integrative
rule pick cards from doors C and D, which are advantageous in the long
run.

% problematic aspects of standard analysis
Typical analyses of these data use the last 60 trials in a series of
200 trials.  A silent assumption made in these analyses is
that behavior has stabilized after 140 learning trials; this could
very well be wrong and it is highly likely that there are individual
differences in this learning process.

% proposed analysis here
Here, we analyze IGT learning data from a single participant with the aim to
establish 1) at which trial learning stops, i.e., at which
trial behavior has stabilized, and 2) what strategies participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
various participants.  The responses were fitted with multinomial
logistic models for the 4 possible choices that participants make in
this task.  There were no covariates on any of the parameters.  The
only constraint imposed on the models' parameters was that
there were designated begin and end states.  This means that the
initial state probabilities were fixed to one for the first state and
to zero for the remaining states.  For the transition matrix this means
that the final state was an absorbing state: the probability of
transitioning out of that state is zero.  This was done to ensure that
there would be a final state in which participants end which provides
the possibility of immediately seeing participants' behavior at the
end of the task.  Models were selected using the Akaike Information
Criterion (AIC; Akaike, 1973).

@@

%We chose to analyze two participants that used different strategies at
%the end of the task as analyzed in the manner proposed by
%\citet{Huizenga2007}.  The first participants' data 
The data were best 
described by a 4-state model.  This model's transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix the final state is
absorbing.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~\ref{tab:igt4} below.

@@

The states have clear interpretations. States 3 and 4 are both
dominated by C responses, and state 4 has a low probability of B
responses. States 1 and 2 are dominated C and D and A and B responses 
respectively. To get a clearer interpretation of these states it is
necessary to consider when they are visited during the learning
process. The Viterbi algorithm (REFERENCE) provides us with the
posterior state sequence, i.e., the sequence of states that the
participant is in at each trial. This sequence is depicted in
Figure~\ref{fig:post4}.

@@

At the start of training there are many switches between states 1 and 
2, indicating that behavior is mostly random choices between all four 
categories, albeit with a slight preference for B and D over A and C
choices. This preference indicates a focus on the frequency of loss
associated with each of the doors as B and D have low frequency of
loss. After this, there is a period of stable state 2 behavior
associated with responses A and B. After that there is stable state 3 
behavior, consisting only of C responses, a short transitional period 
and then state 4 behavior consisting of 94 \% C choices and some B
choices, i.e., almost optimal behavior. The choice proportions and the
model predicted choice proportions are depicted in
Figure~\ref{fig:igt4}.

@@

The initial preference for B and D choices confirms the theory
expressed in \citet{Huizenga2007} that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices represents one of the optimal strategies, which consists of
both C and D responses. Note that both C and D responses generate equal
profits in the long run.

@@

There are different accounts of probabilistic category learning 
\cite<see e.g.>[for an overview]{Ashby2005}.
According to instance or exemplar learning theories, participants
learn by storing each encountered cue-outcome pair.  When presented
with a probe cue pattern, a response is made by retrieving these 
exemplars are retrieved from memory and weighting them according 
to their similarity to the probe cue pattern. According to associative theories,
participants learn by gradually associating the individual
cues (or cue patterns in configural learning) to the outcomes. Finally, in
rule-learning, participants are taken to extract rules by which to
categorize the different cue patterns.  \cite{Gluck2002} proposed a
number of such rules (or strategies). The rules differ in the way
the cues are combined into a response. A main difference is whether responses are
based on the presence/absence of a single cue, or whether information 
from multiple cues is integrated. \citeauthor{Gluck2002}
formulated all strategies in a deterministic and optimal manner (e.g., 
the multi-cue strategy corresponded to giving the optimal response to
each cue pattern). \citet{Meeter2006}  allowed for probabilistic
responding (a small probability of responding non-optimally) as well
as switches between strategies during learning. However, both strategies 
and response probabilities remained predefined. Alternative non-strategy based 
analyses of the WPT \cite{Lagnado2006,Speekenbrink2008} have estimated 
response strategies through logistic regression, allowing the regression 
coefficients to change smoothly over time.

%associative, rule-based.
Here, we combine the regression and strategy-based approaches, and analyse 
the behavior of a single individual performing the
WPT for 200 trials.  We are particularly interested
in evidence for strategy switching and whether a DMM can recover
strategies in accordance with \citeA{Gluck2002}. We chose to analyse the 
``average'' participant (the participant with performance closest to 
the group average) in a large unpublished dataset. We let each state 
be characterized by a GLM with a Binomial distributed response and 
logistic link function (i.e., a logistic regression model). The regression
coefficients of a model relating the cues to the state of the weather are 
given in column ```validity'' of Table~\ref{tab:WPT}. Note that indentical
coefficients for the model relating the cues to responses would indicate 
probability matching. A maximizing strategy is indicated by more
extreme regression coefficients.

As we fitted a DMM to the data of a single subject, it was necessary
to place some constraints on the model.  Specifically, we constrained
the state transitions to be in a ``left-right'' format (states can
only proceed to the immediately adjacent state and never back, and
must start in the initial state).  We fitted a single, two and three
state model to the data.  This showed that a two state model was
better than a single or three state model.

\begin{table}
\caption{Estimates for the weather prediction task}
\label{tab:WPT}
\begin{tabular}{lcccccccccc} \hline
 & & \multicolumn{1}{c}{} && \multicolumn{1}{c}{1 state} & & \multicolumn{2}{c}{2 state} &&
 \multicolumn{2}{c}{2 state (constr.)} \\ \cline{5-5} \cline{7-8} \cline{10-11}
parameter && validity && $S_1$ & & $S_1$ & $S_2$ & & $S_1$ & $S_2$ \\ \hline
(intercept) && 0 && -0.69 & & -2.73 & 0.88 & & -1.24 & 0 \\
cue 1 && 2.10 && 1.69 && 2.12 & 1.60 && 1.65 & 1.97 \\
cue 2 && 0.58 && 1.12 && 0.97 & 1.63 && 0 & 1.92 \\
cue 3 && -0.58 && -0.49 && 0.91 & -2.03 && 0 & -1.58 \\
cue 4 && -2.10 && -1.32 && 0.69 & -3.16 && 0 & -2.67 \\ \hline
 & & & & \multicolumn{1}{c}{AIC=204.47} & & \multicolumn{2}{c}{AIC=187.50} &&
 \multicolumn{2}{c}{AIC=185.24}
\end{tabular}
\end{table}

Investigation of the parameter estimates (see Table~\ref{tab:WPT})
showed that the regresion coefficient for the first cue was of much
larger magnitude than that of the other cues. As such, the first
state seems representative of single cue strategy.  Alternatively, as
all regression coefficients are positive, it could indicate a
``counting'' heuristic, where the propensity of ``sun'' responses
increases when more cues are present (regardless of which cues they
are).  However, in that case, we would expect the regression
coefficients to be of roughly identical magnitude. The second state
seemed to represent a multi-cue strategy, as all cues had regression
coefficients of reasonable magnitude in the direction of the objective
cue validities.

To reduce the degrees of freedom and improve parameter estimates, we
implemented constraints to force state 1 into a single cue strategy
(fixing the coefficients of the remaining three cues to 0) and state 2
in a multi-cue strategy (forcing the intercept to 0).  These
restrictions resulted in a better AIC value of AIC=185.24 (df=7).
Interestingly, the single cue strategy was somewhat different than
described by \citet{Gluck2002}.  Parameter estimates indicated relatively
more consistent predictions of ``rain'' in the absence of cue 1
($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun''
in the presence of cue 1 ($Pr(\text{sun}) = 0.60$).  The cue weights
of the multi-cue strategy were in the direction of the optimal
weights.  The Viterbi state sequence indicated that the participant
used the single cue strategy for the first 60 trials, and then
switched to the multi-cue strategy.
